\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-07-22 analysis/paper/paper.Rmd

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- FALSE
eval_cs4 <- TRUE
eval_colophon <- TRUE

@@

```{r stanford-hazard, eval = eval_cs4, echo=TRUE}
# Compute hazards---
# First, create a list of time points for both exposure status
hazard_data <- expand.grid(
  exposure = c(0, 1),
  futime = seq(0, 1000,
    length.out = 100
  )
)

@@

```{r stanford-risk, eval = eval_cs4}
# Compute absolute risk curves
newdata <- data.frame(exposure = c(0, 1))
absrisk <- absoluteRisk(fit4,
  newdata = newdata,
  time = seq(0, 1000, length.out = 100)
)

@@

We can also plot the hazard ratio:
```{r eval = eval_cs4}
newtime <- quantile(fit4[["originalData"]][[fit4[["timeVar"]]]],
                    probs = seq(0.01, 0.99, 0.01))

@@

plot(fit4,
     type = "hr",
     newdata = newdata,
     var = "exposure",
     increment = 1,
     xvar = "futime",
     ci = FALSE,
     rug = TRUE)
```

@@

In this article, we presented the \proglang{R} package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling. We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through four case studies.

As a methodological framework, case-base sampling is very flexible. This flexibility has been explored before in the literature: for example, Saarela and Hanley [-@saarela2015case] used case-base sampling to model a time-dependent exposure variable in a vaccine safety study. As another example, Saarela and Arjas [-@saarela2015non] combined case-base sampling and a Bayesian non-parametric framework to compute individualized risk assessments for chronic diseases. In the case studies above, we explored this flexibility along two fronts. On the one hand, we showed how splines could be used as part of the linear predictor to model the effect of time on the hazard. This strategy yielded estimates of the survival function that were qualitatively similar to semiparametric estimates derived from Cox regression; however, case-base sampling led to estimates of the survival function that *vary smoothly in time*. On the other hand, we also displayed the flexibility of case-base sampling by showing how it could be combined with penalized logistic regression to perform variable selection. Even though we did not illustrate it in this article, case-base sampling can also be combined with the framework of *generalized additive models*. This functionality has been implemented in our package \pkg{casebase}. Similarly, case-base sampling can also be combined with quasi-likelihood estimation to fit survival models that can account for the presence of over-dispersion. All of these examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.

@@

Future work will look at some of the methodological extensions of case-base sampling. First, to assess the quality of the model fit, we want to study the properties of the residuals (e.g. Cox-Snell, martingale). More work needs to be done to understand these residuals in the context of the partial likelihood underlying case-base sampling. The resulting diagnostic tools would then be integrated in this package. Also, we are interested in extending case-base sampling to account for interval censoring. This type of censoring is very common in longitudinal studies, and many packages (e.g. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2}) provide functions to account for it. Again, we hope to include any resulting methodology as part of this package.

>>>>

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- TRUE
eval_colophon <- TRUE

@@

```{r stanford-hazard, eval = eval_cs4, echo=TRUE}
# Compute hazards---
# First, create a list of time points for both exposure status
hazard_data <- expand.grid(
  exposure = c(0, 1),
  futime = seq(0, 425,
    length.out = 100
  )
)

@@

```{r stanford-risk, eval = eval_cs4}
# Compute absolute risk curves
newdata <- data.frame(exposure = c(0, 1))
absrisk <- absoluteRisk(fit4,
  newdata = newdata,
  time = seq(0, 425, length.out = 100)
)

@@

We can also plot the hazard ratio:
```{r eval = eval_cs4}
newtime <- seq(0, 425, length.out = 100)

@@

plot(fit4,
     type = "hr",
     newdata = newdata,
     var = "exposure",
     increment = 1,
     xvar = "futime",
     ci = FALSE,
     rug = TRUE)
abline(h = 1, lty = 2)
```

@@

In this article, we presented the \proglang{R} package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling. Our package also provide several functions to produce graphical summaries of the data and the results. We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through four case studies.

As a methodological framework, case-base sampling is very flexible. This flexibility has been explored before in the literature: for example, Saarela and Hanley [-@saarela2015case] used case-base sampling to model a time-dependent exposure variable in a vaccine safety study. As another example, Saarela and Arjas [-@saarela2015non] combined case-base sampling and a Bayesian non-parametric framework to compute individualized risk assessments for chronic diseases. In the case studies above, we further explored this flexibility along two fronts. On the one hand, we showed how splines could be used as part of the linear predictor to model the effect of time on the hazard. This strategy yielded estimates of the survival function that were qualitatively similar to semiparametric estimates derived from Cox regression; however, case-base sampling led to estimates of the survival function that *vary smoothly in time* and are thus easier to interpret. On the other hand, we also displayed the flexibility of case-base sampling by showing how it could be combined with penalized logistic regression to perform variable selection. Even though we did not illustrate it in this article, case-base sampling can also be combined with the framework of *generalized additive models*. This functionality has already been implemented in the package. Similarly, case-base sampling can be combined with quasi-likelihood estimation to fit survival models that can account for the presence of over-dispersion. All of these examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.

@@

Future work will look at some of the methodological extensions of case-base sampling. First, to assess the quality of the model fit, we would like to study the properties of the residuals (e.g. Cox-Snell, martingale). More work needs to be done to understand these residuals in the context of the partial likelihood underlying case-base sampling. The resulting diagnostic tools could then be integrated in this package. Also, we are interested in extending case-base sampling to account for interval censoring. This type of censoring is very common in longitudinal studies, and many packages (e.g. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2}) provide functions to account for it. Again, we hope to include any resulting methodology as part of this package.
